---
title: "Add classification parameter to a FCS file"
author: "Anne-Maud Ferreira"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      results = FALSE,
                      warning = FALSE, 
                      messsage = FALSE,
                      cache = FALSE)
```

The goal is to add a classification parameter to a FCS file.

# Environment set up

Load libraries:

```{r load_libraries}
suppressPackageStartupMessages({
  library(flowCore)
  library(tidyverse) 
  library(BiocGenerics) 
})
```

# Input data

This part need to be adapted with the user's data.

## Load FCS file

The users provides the path to the FCS file.

```{r read_fcsdata}
# Path # TO ADAPT
fcs_path <- "/g/bdimsort/220228_Salmonella/20220228_2017_B2_wt_MOI20_50k/B2_wt_MOI20_50k.fcs"
# Read the data:
original <- read.FCS(fcs_path)
```

## Manual classification directory

The users provides the path to the directory containing the classification.

1. List the files in the directory:

```{r list_files_class}
# Path to the directory # TO ADAPT
dir_path <- "/g/bdimsort/220228_Salmonella/20220228_2017_B2_wt_MOI20_50k/classification_camilla/"
# List files
v_all_files <- list.files(dir_path, recursive = TRUE)
```

2. Split directory names to create a table of the manual classification:

```{r split_directory_names}
# Directory names
tb_classified_files <- data.frame("path" = v_all_files, 
                                  "directory_name" = dirname(v_all_files),
                                  "file_name" = basename(v_all_files))
```

3. Extract the ID code: this 8 digits code correspond to the row of the FCS file, i.e.
each row is a cell in the FCS object

```{r extract_ID}
# Extract the ID code from file name # TO ADAPT with file name if the barcode is not located at the end of the file name
tb_classified_files$barcode <- sub('.*(\\d{8}).*', '\\1', 
                                   tb_classified_files$file_name)
# Extract row number for the barcode
tb_classified_files$row <- sub("0+(.*)", "\\1", tb_classified_files$barcode)
```

4. Create classification number to add to the FCS file as it must to be numeric values:

```{r table_classification}
# Convert class to a number
tb_classified_files$manual_classification <- 
  as.numeric(as.factor(tb_classified_files$directory_name))
```

# Create the classification parameter

Create a vector that will be added to the FCS file:

```{r create_col}
# Extract dimensions of the exprs values
dim_original <- dim(exprs(original))
# Create a vector of zeros
# zero = events in the exprs table that have not been classified
v_col <- rep(0, dim_original[1])
# Modify vector with the manual classification
v_col[as.numeric(tb_classified_files$row)] <- tb_classified_files$manual_classification
```

# Add the new classification parameter

Add the column to the FCS file (based on code provided [here](https://stackoverflow.com/questions/53269975/using-r-and-flowcores-write-fcs-to-append-a-parameter)):

```{r add_column}
## Let's create a new parameter as an AnnotatedDataFrame by copying the first parameter from the original flowFrame
new_p <- parameters(original)[1,]

## Now, let's change it's name from $P1 to $Px (whatever the next new number is)
new_p_number <- as.integer(dim(original)[2]+1)
rownames(new_p) <- c(paste0("$P", new_p_number))

## Now, let's combine the original parameter with the new parameter 
## BiocGenerics for the combine function
allPars <- BiocGenerics::combine(parameters(original), new_p)

## Fix the name and description of the newly added parameter
new_p_name <- "manual_classification" # TO ADAPT if needed
allPars@data$name[new_p_number] <- new_p_name
allPars@data$desc[new_p_number] <- new_p_name

## Check that allPars contains what it should
allPars@data

## Let's get our cluster ID into a single column matrix
## Using random numbers here; replace with your own code as appropriate
orig_col_names <- dimnames(original@exprs)[[2]] 
class_ids <- as.matrix(v_col, ncol=1)
new_exprs <- cbind(original@exprs, class_ids)
new_par_col_name <- setNames(new_p_name,
                             paste0("$P",as.character(new_p_number),"N"))
dimnames(new_exprs)[[2]] <- c(orig_col_names, new_par_col_name)

## Now, let's get all the original keywords and let's add to it
new_kw <- original@description
new_kw["$PAR"] <- as.character(new_p_number)
new_kw[paste0("$P",as.character(new_p_number),"N")] <- new_p_name
new_kw[paste0("$P",as.character(new_p_number),"S")] <- new_p_name
new_kw[paste0("$P",as.character(new_p_number),"E")] <- "0,0"
new_kw[paste0("$P",as.character(new_p_number),"G")] <- "1"
new_kw[paste0("$P",as.character(new_p_number),"B")] <- new_kw["$P1B"]
new_kw[paste0("$P",as.character(new_p_number),"R")] <- new_kw["$P1R"]
new_kw[paste0("flowCore_$P",as.character(new_p_number),"Rmin")] <- new_kw["flowCore_$P1Rmin"]
new_kw[paste0("flowCore_$P",as.character(new_p_number),"Rmax")] <- new_kw["flowCore_$P1Rmax"]

## Now, let's just combine it into a new flowFrame
new_fcs <- new("flowFrame", exprs=new_exprs, parameters=allPars, description=new_kw)
```

# Output files

The user provides the path where to save the output file:

```{r path_output_file}
# Path # TO ADAPT
output_path <- "/g/steinmetz/project/ICS/results/"
```

## Classification code

Create a table with the classification code that has been used:

```{r table_classification_code}
# Summarize the manual classification
table_manualclassification <- unique(tb_classified_files[,c("directory_name", 
                              "manual_classification")])
# Add zero if an image was not classify
table_classcode <- data.frame("groups" = c(table_manualclassification$directory_name, "no_classification"),
                              "manual_classification" = c(table_manualclassification$manual_classification, 0))
```

Save the table containing the used classification code:

```{r save_classification_code}
# Save table
write.table(table_classcode, 
            file = paste0(output_path, "table_classcode.txt"),
            col.names = TRUE,
            row.names = FALSE)
```

## New FCS file

Save the new FCS file:

```{r save_FCS_file}
# Create a new FCS name combining the original name and the classification directory
## FCS file name
fcs_name <- basename(fcs_path)
## Substract the .fcs extension
fcs_name_woext <- sub("(*).fcs", "\\1", fcs_name)
## Directory name
dir_name <- basename(dir_path)
## New FCS file name
new_fcs_filename <- paste0(fcs_name_woext, "_", dir_name, ".fcs")

# Save using the regular write.FCS from flowCore to save the new FCS file
write.FCS(new_fcs, 
          filename = paste0(output_path, new_fcs_filename), 
          delimiter = "#")
## This new file should now be readable nicely R or any other software.
```

